---
title: "Missing Data Techniques"
output: html_notebook
---
```{r setup, include=FALSE}
# library
library("readr")
library("rstanarm")
library("ggplot2")
library("bayesplot")
library("mice")
library("dplyr")
library(VIM)
library(mi)
```
## For this assignment I am using the Motor Vehicle Collision – Crashes on NYC Open Data. I downloaded it from this url: [https://data.cityofnewyork.us/Public-Safety/Motor- Vehicle-Collisions-Crashes/h9gi-nx95] filtering for data collected between 01/01/2021 through 02/01/2022. I plan to conduct linear regression on the data. With, “number of persons injured”, as the dependent variable and using the borough as the independent variable.

```{r}
# loading in motor vehicle dataset
vehicle_df <-  read_csv("Desktop/ASSR/Missing Data/Motor_Vehicle_Collisions_-_Crashes.csv", 
    col_types = cols(LOCATION = col_skip(), 
        `CONTRIBUTING FACTOR VEHICLE 1` = col_skip(),
        `CONTRIBUTING FACTOR VEHICLE 2` = col_skip(),
        `CONTRIBUTING FACTOR VEHICLE 3` = col_skip(), 
        `CONTRIBUTING FACTOR VEHICLE 4` = col_skip(), 
        `CONTRIBUTING FACTOR VEHICLE 5` = col_skip(), 
        `CROSS STREET NAME` = col_skip(),
        `OFF STREET NAME` = col_skip(),
        `VEHICLE TYPE CODE 1` = col_skip(), 
        `VEHICLE TYPE CODE 2` = col_skip(),
        `VEHICLE TYPE CODE 3` = col_skip(),
        `VEHICLE TYPE CODE 4` = col_skip(), 
        `VEHICLE TYPE CODE 5` = col_skip()))
head(vehicle_df)
# saving modified df for tableau summary statistics
# write.csv(vehicle_df,"~/Desktop/ASSR/Missing Data/vehicle_df.csv", row.names = FALSE)
```

# 1. Provide some plots and summary statistics, like percent missing per variable, percent complete cases, and so on.
```{r}
 # Check % missing for 
n = nrow(vehicle_df)
sum(is.na(vehicle_df))/n
```

```{r}
# table of missing data by column
colMeans(apply(vehicle_df, 2, is.na))
```
```{r}
# counting the number of missing values for each column
colSums(is.na(vehicle_df))
```

# After each of the following tasks, you need to implement the analysis you have in mind and report the results/estimates.
-------------------------------------------------------------------------------------------
# 2. Listwise deletion
```{r}
# performing listwise deletion to get rid of cases with missing categorical variables.
sum(is.na(vehicle_df)) # checking how many na values are there prior to listwise deletion
likewise_deletion_vehicle_df <- na.omit(vehicle_df) # preforming listwise deletion
sum(is.na(likewise_deletion_vehicle_df)) # checking how many na values are there after listwise deletion
```
```{r}
# displaying the full dataset after removing all possible na values
View(likewise_deletion_vehicle_df)
```
```{r}
# conducting linear regression on the likewise deletion data, with, “number of persons injured”, as the dependent variable and using "borough" as the independent variable.
fit2 <- lm(data = likewise_deletion_vehicle_df, `NUMBER OF PERSONS INJURED` ~ BOROUGH)
summary(fit2)
```
```{r}
#histogram of residuals
ggplot(data = likewise_deletion_vehicle_df, aes(x = fit2$residuals)) +
    geom_histogram(bins = 20, fill = 'orange', color = 'black') +
    labs(title = 'Likewise Deletion Residuals', x = 'Residuals', y = 'Frequency', caption = "Source: Vehicle Collision Data") +
  theme_classic() +
  theme(plot.title = element_text(color = "black", size = 12, hjust = 0.5 ), plot.caption = element_text(face = "italic"))

```


-------------------------------------------------------------------------------------------
# 3. Mean/mode imputation.
## Mean Imputation on quantitative variables:
```{r}
# creating a function to take the mean of a column based on the example from class
mean.imputation <- function(a) {
  missing <- is.na(a)
  a.obs <- a[!missing]
  imputed <- a
  imputed[missing] <- mean(a.obs)
  return (imputed)
}

# creating a new dataframe to store the modifications
mean_mode_vehicle_df <- vehicle_df

# Preforming mean imputation on, "ZIP CODE" variable
mean_mode_vehicle_df$`ZIP CODE` <- mean.imputation(mean_mode_vehicle_df$`ZIP CODE`)
sum(is.na(mean_mode_vehicle_df$`ZIP CODE`))

# Preforming mean imputation on, "NUMBER OF PERSONS INJURED" variable
mean_mode_vehicle_df$`NUMBER OF PERSONS INJURED` <- mean.imputation(mean_mode_vehicle_df$`NUMBER OF PERSONS INJURED`)
sum(is.na(mean_mode_vehicle_df$`NUMBER OF PERSONS INJURED`))

# Preforming mean imputation on, "LATITUDE" variable
mean_mode_vehicle_df$LATITUDE <- mean.imputation(mean_mode_vehicle_df$LATITUDE)
sum(is.na(mean_mode_vehicle_df$LATITUDE))

# Preforming mean imputation on, "LONGITUDE" variable
mean_mode_vehicle_df$LONGITUDE <- mean.imputation(mean_mode_vehicle_df$LONGITUDE)
sum(is.na(mean_mode_vehicle_df$LONGITUDE))
```

# 2. performing mode imputation on categorical variables
## The categorical variables, "on street name", "contributing factor vehicle 1", "contributing factor vehicle 2", "vehicle type code 1", "vehicle type code 2", "borough have na values.
```{r}
# creating a function that finds the mode of a variable and then replaces the missing data with that mode
mode = function(x) {
ta = table(x)
tam = max(ta)
if (all(ta == tam))
  mod = NA else
    mod = names(ta)[ta == tam] 
return(mod)
}

mode.imp <- function (a) {
missing <- is.na(a)
a.obs <- a[!missing]
imputed <- a
imputed[missing] <- mode(a.obs) 
return(imputed)
}

# Preforming mode imputation on, "BOROUGH" variable
mean_mode_vehicle_df$BOROUGH <- mode.imp(mean_mode_vehicle_df$BOROUGH)
sum(is.na(mean_mode_vehicle_df$BOROUGH))

# Preforming mode imputation on, "ON STREET NAME" variable
mean_mode_vehicle_df$`ON STREET NAME`<- mode.imp(mean_mode_vehicle_df$`ON STREET NAME`)
sum(is.na(mean_mode_vehicle_df$`ON STREET NAME`))
```
```{r}
# checking that all missing values have been replaced with either mean or mode imputation
sum(is.na(mean_mode_vehicle_df))
```

```{r}
# mean mode data analysis

head(mean_mode_vehicle_df)
# conducting linear regression on the modified data, with, “number of persons injured”, as the dependent variable and using "borough" as the independent variable.
fit3 <- lm(data = mean_mode_vehicle_df, `NUMBER OF PERSONS INJURED` ~ BOROUGH)
summary(fit3)
```
```{r}
#histogram of residuals
ggplot(data = mean_mode_vehicle_df, aes(x = fit3$residuals)) +
    geom_histogram(bins = 20, fill = 'orange', color = 'black') +
    labs(title = 'Mean and Mode Imputation Residuals', x = 'Residuals', y = 'Frequency', caption = "Source: Vehicle Collision Data") +
  theme_classic() +
  theme(plot.title = element_text(color = "black", size = 12, hjust = 0.5 ), plot.caption = element_text(face = "italic"))

```

-------------------------------------------------------------------------------------------
# 4. Random imputation.
```{r}
# creating dataframe with only quantitative variables for regression imputation
quant_df <- data.frame('ZIP CODE' = vehicle_df$`ZIP CODE`,'LATITUDE' = vehicle_df$LATITUDE, 'NUMBER OF PERSONS INJURED' = vehicle_df$`NUMBER OF PERSONS INJURED`, 'NUMBER OF PERSONS KILLED' = vehicle_df$`NUMBER OF PERSONS KILLED`,'NUMBER OF PEDESTRIANS INJURED' = vehicle_df$`NUMBER OF PEDESTRIANS INJURED`, 'NUMBER OF PEDESTRIANS KILLED' = vehicle_df$`NUMBER OF PEDESTRIANS KILLED`, 'NUMBER OF CYCLIST INJURED' = vehicle_df$`NUMBER OF CYCLIST INJURED`, 'NUMBER OF CYCLIST KILLED' = vehicle_df$`NUMBER OF CYCLIST KILLED`, 'NUMBER OF MOTORIST INJURED' = vehicle_df$`NUMBER OF MOTORIST INJURED`,  'NUMBER OF MOTORIST KILLED' = vehicle_df$`NUMBER OF MOTORIST KILLED`)
head(quant_df)
```

```{r}
# checking to see how many na values in the quant df
sum(is.na(quant_df$ZIP.CODE))
sum(is.na(quant_df$LATITUDE))
sum(is.na(quant_df$NUMBER.OF.PERSONS.INJURED))
sum(is.na(quant_df$NUMBER.OF.PEDESTRIANS.KILLED))
sum(is.na(quant_df$NUMBER.OF.CYCLIST.INJURED))
sum(is.na(quant_df$NUMBER.OF.CYCLIST.KILLED))
sum(is.na(quant_df$NUMBER.OF.MOTORIST.INJURED))
sum(is.na(quant_df$NUMBER.OF.MOTORIST.KILLED))
```

```{r}
# random imputation on the quantitative variables 
random_imp_vehicle_df.complete = complete(mice(quant_df, method = "sample", m = 1, maxit = 1))
```
```{r}
# checking the number of na values in the new df
sum(is.na(random_imp_vehicle_df.complete$ZIP.CODE))
sum(is.na(random_imp_vehicle_df.complete$LATITUDE))
sum(is.na(random_imp_vehicle_df.complete$NUMBER.OF.PERSONS.INJURED))
sum(is.na(random_imp_vehicle_df.complete$NUMBER.OF.PERSONS.KILLED))
sum(is.na(random_imp_vehicle_df.complete$NUMBER.OF.PEDESTRIANS.INJURED))
sum(is.na(random_imp_vehicle_df.complete$NUMBER.OF.PEDESTRIANS.KILLED))
sum(is.na(random_imp_vehicle_df.complete$NUMBER.OF.CYCLIST.INJURED))
sum(is.na(random_imp_vehicle_df.complete$NUMBER.OF.CYCLIST.KILLED))
sum(is.na(random_imp_vehicle_df.complete$NUMBER.OF.MOTORIST.INJURED))
sum(is.na(random_imp_vehicle_df.complete$NUMBER.OF.MOTORIST.KILLED))
```

```{r}
# replacing the columns with missing values as part of the regular dataframe
random_imp_vehicle_df <- vehicle_df
random_imp_vehicle_df$'ZIP CODE' <- random_imp_vehicle_df.complete$ZIP.CODE
random_imp_vehicle_df$LATITUDE <- random_imp_vehicle_df.complete$LATITUDE
random_imp_vehicle_df$`NUMBER OF PERSONS INJURED` <- random_imp_vehicle_df.complete$NUMBER.OF.PERSONS.INJURED
random_imp_vehicle_df$`NUMBER OF PEDESTRIANS KILLED` <- random_imp_vehicle_df.complete$NUMBER.OF.PERSONS.KILLED
random_imp_vehicle_df$`NUMBER OF CYCLIST INJURED` <- random_imp_vehicle_df.complete$NUMBER.OF.CYCLIST.INJURED
random_imp_vehicle_df$`NUMBER OF CYCLIST KILLED` <- random_imp_vehicle_df.complete$NUMBER.OF.CYCLIST.KILLED
random_imp_vehicle_df$`NUMBER OF MOTORIST INJURED` <- random_imp_vehicle_df.complete$NUMBER.OF.MOTORIST.INJURED
random_imp_vehicle_df$`NUMBER OF MOTORIST KILLED` <- random_imp_vehicle_df.complete$NUMBER.OF.MOTORIST.KILLED
```


```{r}
# regression data analysis
head(random_imp_vehicle_df)
# conducting linear regression on the modified data, with, “number of persons injured”, as the dependent variable and using "borough" as the independent variable.
fit6 <- lm(data = random_imp_vehicle_df, `NUMBER OF PERSONS INJURED` ~ BOROUGH)
summary(fit6)
```
-------------------------------------------------------------------------------------------
# 5. Cannot use LVCF method because the data is not longitudial data.
-------------------------------------------------------------------------------------------
#6. Hotdecking (nearest neighbor) with VIM package.
```{r}
# Nearest neighbor hotdecking
hotdeck_vehicle_df <- vehicle_df
sum(is.na(hotdeck_vehicle_df))

# Using Manhattan distance and variance as weight to rescale variables:
hotdeck_vehicle_df <- hotdeck(data = hotdeck_vehicle_df, 
        variable = c("BOROUGH", "LATITUDE", "ZIP CODE", "LONGITUDE", "ON STREET NAME", "NUMBER OF PERSONS INJURED"),
        domain_var = "CRASH TIME")
sum(is.na(hotdeck_vehicle_df))
```
```{r}
fit6.5 <- lm(data = hotdeck_vehicle_df, `NUMBER OF PERSONS INJURED` ~ BOROUGH)
summary(fit6.5)
```
```{r}
#histogram of residuals
ggplot(data = hotdeck_vehicle_df, aes(x = fit6.5$residuals)) +
    geom_histogram(bins = 20, fill = 'orange', color = 'black') +
    labs(title = 'Hotdeck Residuals', x = 'Residuals', y = 'Frequency', caption = "Source: Vehicle Collision Data") +
  theme_classic() +
  theme(plot.title = element_text(color = "black", size = 12, hjust = 0.5 ), plot.caption = element_text(face = "italic"))
```

-------------------------------------------------------------------------------------------
#7. Regression imputation. (Note you might have to use logistic or multinomial models, depending on what type of variable you impute values for.)
```{r}
# checking if their are na values in the new dataset prior to imputation
sum(is.na(quant_df$ZIP.CODE))
sum(is.na(quant_df$LATITUDE))
sum(is.na(quant_df$NUMBER.OF.PERSONS.INJURED))
sum(is.na(quant_df$NUMBER.OF.PEDESTRIANS.KILLED))
sum(is.na(quant_df$NUMBER.OF.CYCLIST.INJURED))
sum(is.na(quant_df$NUMBER.OF.CYCLIST.KILLED))
sum(is.na(quant_df$NUMBER.OF.MOTORIST.INJURED))
sum(is.na(quant_df$NUMBER.OF.MOTORIST.KILLED))
```
```{r}
# Preforming regression imputation on, the quantitative variables
imp <- mice(quant_df, method = "norm.predict", m = 1)
data_imp <- complete(imp)
head(data_imp)
```

```{r}
sum(is.na(data_imp$ZIP.CODE))
sum(is.na(data_imp$LATITUDE))
sum(is.na(data_imp$NUMBER.OF.PERSONS.INJURED))
sum(is.na(data_imp$NUMBER.OF.PERSONS.KILLED))
sum(is.na(data_imp$NUMBER.OF.PEDESTRIANS.INJURED))
sum(is.na(data_imp$NUMBER.OF.PEDESTRIANS.KILLED))
sum(is.na(data_imp$NUMBER.OF.CYCLIST.INJURED))
sum(is.na(data_imp$NUMBER.OF.CYCLIST.KILLED))
sum(is.na(data_imp$NUMBER.OF.MOTORIST.INJURED))
sum(is.na(data_imp$NUMBER.OF.MOTORIST.KILLED))
```
```{r}
# making the column with missing values a part of the regular dataframe
regression_df <- vehicle_df
regression_df$'ZIP CODE' <- data_imp$ZIP.CODE
regression_df$LATITUDE <- data_imp$LATITUDE
regression_df$`NUMBER OF PERSONS INJURED` <- data_imp$NUMBER.OF.PERSONS.INJURED
regression_df$`NUMBER OF PEDESTRIANS KILLED` <- data_imp$NUMBER.OF.PERSONS.KILLED
regression_df$`NUMBER OF CYCLIST INJURED` <- data_imp$NUMBER.OF.CYCLIST.INJURED
regression_df$`NUMBER OF CYCLIST KILLED` <- data_imp$NUMBER.OF.CYCLIST.KILLED
regression_df$`NUMBER OF MOTORIST INJURED` <- data_imp$NUMBER.OF.MOTORIST.INJURED
regression_df$`NUMBER OF MOTORIST KILLED` <- data_imp$NUMBER.OF.MOTORIST.KILLED

sum(is.na(regression_df$'ZIP CODE'))
sum(is.na(regression_df$LATITUDE))
sum(is.na(regression_df$`NUMBER OF PERSONS INJURED`))
sum(is.na(regression_df$`NUMBER OF PEDESTRIANS KILLED`))
sum(is.na(regression_df$`NUMBER OF CYCLIST INJURED`))
sum(is.na(regression_df$`NUMBER OF CYCLIST KILLED`))
sum(is.na(regression_df$`NUMBER OF MOTORIST INJURED`))
sum(is.na(regression_df$`NUMBER OF MOTORIST KILLED`))

head(regression_df)
```

```{r}
# regression data analysis
head(regression_df)
# conducting linear regression on the modified data, with, “number of persons injured”, as the dependent variable and using "borough" as the independent variable.
fit7 <- lm(data = regression_df, `NUMBER OF PERSONS INJURED` ~ BOROUGH)
summary(fit7)
```
```{r}
#histogram of residuals
ggplot(data = regression_df, aes(x = fit7$residuals)) +
    geom_histogram(bins = 20, fill = 'orange', color = 'black') +
    labs(title = 'Regression Imputation Residuals', x = 'Residuals', y = 'Frequency', caption = "Source: Vehicle Collision Data") +
  theme_classic() +
  theme(plot.title = element_text(color = "black", size = 12, hjust = 0.5 ), plot.caption = element_text(face = "italic"))
```
-------------------------------------------------------------------------------------------
# 8. Regression imputation with noise on all variables (numerical, dichotomous and multinomial).
```{r}
# using the quantitative variable subset of the dataframe
# checking to see how many na values in the quant df
sum(is.na(quant_df$ZIP.CODE))
sum(is.na(quant_df$LATITUDE))
sum(is.na(quant_df$NUMBER.OF.PERSONS.INJURED))
sum(is.na(quant_df$NUMBER.OF.PEDESTRIANS.KILLED))
sum(is.na(quant_df$NUMBER.OF.CYCLIST.INJURED))
sum(is.na(quant_df$NUMBER.OF.CYCLIST.KILLED))
sum(is.na(quant_df$NUMBER.OF.MOTORIST.INJURED))
sum(is.na(quant_df$NUMBER.OF.MOTORIST.KILLED))
```
```{r}
# Preforming regression imputation on, the quantitative variables using norm.nob to add noise based on the data
imp2 <- mice(quant_df, method = "norm.nob", m = 1) # creating noise
data_imp2 <- complete(imp2)
head(data_imp2)
```
```{r}
sum(is.na(data_imp2$ZIP.CODE))
sum(is.na(data_imp2$LATITUDE))
sum(is.na(data_imp2$NUMBER.OF.PERSONS.INJURED))
sum(is.na(data_imp2$NUMBER.OF.PERSONS.KILLED))
sum(is.na(data_imp2$NUMBER.OF.PEDESTRIANS.INJURED))
sum(is.na(data_imp2$NUMBER.OF.PEDESTRIANS.KILLED))
sum(is.na(data_imp2$NUMBER.OF.CYCLIST.INJURED))
sum(is.na(data_imp2$NUMBER.OF.CYCLIST.KILLED))
sum(is.na(data_imp2$NUMBER.OF.MOTORIST.INJURED))
sum(is.na(data_imp2$NUMBER.OF.MOTORIST.KILLED))
```
```{r}
# making the column with missing values a part of the regular dataframe
regression_df2 <- vehicle_df
regression_df2$'ZIP CODE' <- data_imp2$ZIP.CODE
regression_df2$LATITUDE <- data_imp2$LATITUDE
regression_df2$`NUMBER OF PERSONS INJURED` <- data_imp2$NUMBER.OF.PERSONS.INJURED
regression_df2$`NUMBER OF PEDESTRIANS KILLED` <- data_imp2$NUMBER.OF.PERSONS.KILLED
regression_df2$`NUMBER OF CYCLIST INJURED` <- data_imp2$NUMBER.OF.CYCLIST.INJURED
regression_df2$`NUMBER OF CYCLIST KILLED` <- data_imp2$NUMBER.OF.CYCLIST.KILLED
regression_df2$`NUMBER OF MOTORIST INJURED` <- data_imp2$NUMBER.OF.MOTORIST.INJURED
regression_df2$`NUMBER OF MOTORIST KILLED` <- data_imp2$NUMBER.OF.MOTORIST.KILLED

sum(is.na(regression_df2$'ZIP CODE'))
sum(is.na(regression_df2$LATITUDE))
sum(is.na(regression_df2$`NUMBER OF PERSONS INJURED`))
sum(is.na(regression_df2$`NUMBER OF PEDESTRIANS KILLED`))
sum(is.na(regression_df2$`NUMBER OF CYCLIST INJURED`))
sum(is.na(regression_df2$`NUMBER OF CYCLIST KILLED`))
sum(is.na(regression_df2$`NUMBER OF MOTORIST INJURED`))
sum(is.na(regression_df2$`NUMBER OF MOTORIST KILLED`))

head(regression_df2)
```
```{r}
# regression data analysis
head(regression_df2)
# conducting linear regression on the modified data, with, “number of persons injured”, as the dependent variable and using "borough" as the independent variable.
fit8 <- lm(data = regression_df2, `NUMBER OF PERSONS INJURED` ~ BOROUGH)
summary(fit8)
```

# There are no dichotomous variables in my dataset so I am unable to preform logistic regression on the data.

-------------------------------------------------------------------------------------------
# Multiple imputation with either mice OR mi package:

#9. Load your data into the package. Obtain summary, and graphs of the data and missing patterns.
```{r}
#creating a new dataframe with only the columns of interest because of errors running whenever I attempt to load the original data to a 'mdf' as long as errors with the iterations
vehicle <- vehicle_df[, c("BOROUGH", "NUMBER OF PERSONS INJURED", "NUMBER OF PERSONS KILLED")]
vehicle
```
```{r}
vehicle <- rename(vehicle, num_injured = `NUMBER OF PERSONS INJURED`, num_killed = `NUMBER OF PERSONS KILLED`)
```
```{r}
vehicle <- as.data.frame(vehicle) #converting from tibble to dataframe for mdf conversion
head(vehicle)
```

```{r}
# loading data into the package
mdf <- missing_data.frame(vehicle)
```

```{r}
# Obtaining summary of the data
summary(mdf)
```
```{r}
# creating this in order to display graph without errors
mdf2 <- missing_data.frame(vehicle[1:500,])
```

```{r}
# obtains graphs of the data
image(mdf2)
```
```{r}
# missingness patterns
md.pattern(mdf2)
```

#10. Check your data types and methods and make changes if necessary.
```{r}
show(mdf)
```

#11. Run the mi/mice command and check convergence by traceplots.
```{r}
#mi with 5 chains and 50 iterations
imp.vehicle <- mi(mdf, seed = 124, n.chains = 5, n.iter = 50)
```
```{r}
# check convergence
converged <- mi2BUGS(imp.vehicle)
plot(imp.vehicle)
```

```{r}
# changing the imputation method to Predictive mean matching
mdf <- change(mdf, y = c("BOROUGH", "num_injured","num_killed"), what = "imputation_method", to = "pmm")
```

```{r}
# rerunning mi with 5 chains and 50 iterations
imp.vehicle <- mi(mdf, seed = 124, n.chains = 5, n.iter = 50)
```

```{r}
# checking rhats
mi(imp.vehicle)
```

```{r}
# checking convergence
converged <- mi2BUGS(imp.vehicle)
plot(imp.vehicle)
```

```{r}
#pool data
res.imp.vehicle = mi::pool(num_injured ~ BOROUGH, data = imp.vehicle)
display(res.imp.vehicle)
```
```{r}
vehicle.comp1 = complete(imp.vehicle, m=1)
vehicle.comp1to5 = complete(imp.vehicle, m=5)
```
```{r}
lapply(vehicle.comp1to5, summary)
```
```{r}
vehicle_plots <- as.data.frame(vehicle.comp1to5) 
head(vehicle_plots)
```